---
title: "Validating the event-clock estimator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating the event-clock estimator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
set.seed(42)
```

```{r setup}
library(eventclock)
```

This vignette stress-tests the estimator \(\widehat A = \sum (\Delta L)^2\)
against data simulated from the *exact* transition law of the model
(`ec_simulate_path()`), so every deviation is attributable to the
estimator, not to model error. Helper to turn simulated paths into
`event_prices`:

```{r}
path_to_ep <- function(p) {
  as_event_prices(
    tibble::tibble(time = as.Date("2020-01-01") + p$step, q = p$q),
    clip = c(1e-8, 1 - 1e-8)
  )
}

estimate_A <- function(n_steps, A, q = 0.4, reps = 200, ...) {
  vapply(seq_len(reps), function(i) {
    p <- ec_simulate_path(1, n_steps, q = q, A = A, ...)
    event_clock(path_to_ep(p), methods = "rv")$A
  }, numeric(1))
}
```

## Consistency and finite-sample precision

The estimator is unbiased at every sampling frequency (up to the tiny
\(\sum a_i^2/4\) drift term) and its dispersion shrinks with the number
of observations:

```{r}
grid <- c(15, 30, 60, 120, 250)
sims <- lapply(grid, estimate_A, A = 0.5)
tab <- data.frame(
  n_obs = grid,
  mean = sapply(sims, mean),
  bias_pct = 100 * (sapply(sims, mean) / 0.5 - 1),
  sd = sapply(sims, sd),
  rmse = sapply(sims, function(s) sqrt(mean((s - 0.5)^2)))
)
knitr::kable(tab, digits = 3)
```

Even the paper-sized windows (a one-month window has about 30 daily
observations) are unbiased; the price of short windows is dispersion,
not bias — which is why interval estimates matter.

## Confidence-interval coverage

`event_clock(se = TRUE)` reports quarticity-based standard errors with
log-scale intervals. Their coverage is close to nominal:

```{r}
coverage <- function(conf, reps = 300, n_steps = 60, A = 0.5) {
  hits <- vapply(seq_len(reps), function(i) {
    p <- ec_simulate_path(1, n_steps, q = 0.4, A = A)
    r <- event_clock(path_to_ep(p), methods = "rv", se = TRUE, conf = conf)
    r$ci_lo <= A && A <= r$ci_hi
  }, logical(1))
  mean(hits)
}
data.frame(
  nominal = c(0.90, 0.95),
  empirical = c(coverage(0.90), coverage(0.95))
)
```

## Outcome independence: the measure-robustness property in action

Quadratic variation does not care which outcome eventually realizes —
this is the observable footprint of measure robustness (a change of
measure reweights outcomes; it cannot change the estimate computed from
a given path):

```{r}
p <- ec_simulate_path(400, 60, q = 0.4, A = 0.5)
qv <- vapply(split(p$L, p$path), function(L) sum(diff(L)^2), numeric(1))
J <- vapply(split(p$J, p$path), function(j) j[1], numeric(1))
data.frame(
  outcome = c("J = 1", "J = 0"),
  mean_A_hat = c(mean(qv[J == 1]), mean(qv[J == 0])),
  n_paths = c(sum(J == 1), sum(J == 0))
)
```

Both conditional means sit at the same true \(A = 0.5\): the clock
measures *how much* was learned, not *what* was learned.

## Lumpy information: what the robustness variants actually do

Real event clocks are lumpy — debates and verdicts, not a smooth flow.
With `jump_share`, the simulator concentrates part of \(A\) in a few
jump steps, and the estimator battery reacts exactly as the theory says
it should:

```{r}
lumpy <- lapply(seq_len(200), function(i) {
  p <- ec_simulate_path(1, 60, q = 0.4, A = 0.5,
                        jump_share = 0.5, n_jumps = 1)
  event_clock(path_to_ep(p))[, c("method", "A")]
})
agg <- do.call(rbind, lumpy)
knitr::kable(
  aggregate(A ~ method, data = agg, FUN = mean),
  digits = 3,
  caption = "Mean estimate across 200 lumpy paths (true A = 0.5, half of it in one jump)"
)
```

`rv` recovers total clock time including the jump; `bipower`,
`truncated`, and `largest1` estimate (roughly) the *smooth* component
only — the gap between `rv` and `bipower` is the jumpiness index. None
of these is "wrong"; they answer different questions, and comparing
them is the point.

## Microstructure noise and the snapshot rule

Traded quotes bounce between bid and ask on a coarse tick grid. Additive
noise inflates measured variation at fine sampling — the classic
signature-plot diagnostic, here on simulated data with known truth:

```{r}
p <- ec_simulate_path(1, 2000, q = 0.4, A = 1)
noisy <- ec_ilogit(p$L + rnorm(nrow(p), sd = 0.04))
ep_noisy <- as_event_prices(
  tibble::tibble(time = as.POSIXct("2024-01-01", tz = "UTC") +
                   3600 * p$step, q = noisy),
  clip = c(1e-8, 1 - 1e-8)
)
plot_signature(ec_signature(ep_noisy, max_every = 24))
```

The finest grid overstates the clock (noise variance is counted once
per increment); sparse sampling converges toward the true \(A = 1\).
On real data the same pattern appears — this is the hourly-vs-daily gap
of the 2024 Polymarket series:

```{r}
data(polymarket2024)
plot_signature(ec_signature(as_event_prices(polymarket2024), max_every = 24))
```

which is why the one-snapshot-per-day rule of `pm_daily()` is the
robust default for daily-horizon work.

## Data-quality screening

Before estimating anything, screen the input series:

```{r}
ec_validate(as_event_prices(brexit2016, time = "date", price = "q_leave",
                            market_id = "Brexit: Leave"))
```

`ec_validate()` reports coverage, staleness, tick size, gaps, and
outliers — in the package's flag-don't-drop spirit: nothing is altered,
everything is visible.
