---
title: "Measuring the event clock: Brexit, the 2016 U.S. election, and Polymarket"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measuring the event clock: Brexit, the 2016 U.S. election, and Polymarket}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(eventclock)
```

## The event clock in one paragraph

Ahead of a scheduled event — a referendum, an election, a central-bank
decision — markets learn. The *event clock* measures how much
outcome-relevant information has arrived and when: event-clock time
$A_{t,T}$ is the quadratic variation of the log-odds
$L_t = \mathrm{logit}(q_t)$ of a traded event probability $q_t$,

$$\widehat A_{t,T} \;=\; \sum_{t_i \in [t,T)}
  \bigl[\mathrm{logit}(q_{t_{i+1}}) - \mathrm{logit}(q_{t_i})\bigr]^2 .$$

Because quadratic variation ignores drift and is invariant under
equivalent measure changes, $\widehat A$ needs no assumption about risk
premia, and any constant level distortion of $q$ (discounting, a constant
state-price tilt) drops out. The concept, identification results, and the
empirical applications reproduced below are developed in Hanke, Schadner,
Stöckl, and Weissensteiner (Working Paper), *Learning Before Scheduled
Events: Prediction Markets, State Prices, and Option Valuation*.

## Two classic event windows: Brexit and the 2016 U.S. election

The package ships the two daily event-probability series used in the
working paper, both derived from betting quotes (see `?brexit2016`):

```{r}
data(brexit2016)
data(us2016)

ep_gb <- as_event_prices(brexit2016,
  time = "date", price = "q_leave",
  market_id = "Brexit: Leave", event_date = as.Date("2016-06-23")
)
ep_us <- as_event_prices(us2016,
  time = "date", price = "trump",
  market_id = "US 2016: Trump", event_date = as.Date("2016-11-08")
)

plot_q(ep_gb)
plot_q(ep_us)
```

## Reproducing the headline information-time table

The working paper evaluates the clock from the one-month valuation date
(May 24 / October 10, 2016) over one-week, two-week, and one-month
horizons, with a battery of robustness variants:

```{r}
gb <- event_clock(ep_gb,
  from = as.Date("2016-05-24"),
  to = c(`1W` = as.Date("2016-05-31"), `2W` = as.Date("2016-06-07"),
         `1M` = as.Date("2016-06-23"))
)
us <- event_clock(ep_us,
  from = as.Date("2016-10-10"),
  to = c(`1W` = as.Date("2016-10-17"), `2W` = as.Date("2016-10-24"),
         `1M` = as.Date("2016-11-08"))
)

tab <- rbind(
  cbind(case = "Brexit", as.data.frame(gb[, c("horizon", "method", "A")])),
  cbind(case = "U.S. election", as.data.frame(us[, c("horizon", "method", "A")]))
)
knitr::kable(
  stats::reshape(tab,
    idvar = c("case", "horizon"), timevar = "method", direction = "wide"
  ),
  digits = 3,
  col.names = c("Case", "Horizon", "RV", "Truncated", "Bipower",
                "Largest-1", "Largest-2")
)
```

This reproduces the published table: Brexit $\widehat A$ of 0.064 (1W),
0.166 (2W), and 0.511 (1M) with bipower 0.375; U.S. election 0.015,
0.046, and 0.370 with bipower 0.386 and truncated 0.259.

Two details worth noting:

* **Truncation vs. largest-move exclusion.** The truncation rule (drop
  increments beyond 3 robust standard deviations) binds for the U.S.
  window — where it removes exactly the largest move and matches the
  published 0.259 — but not for the Brexit window, whose largest move
  stays inside the threshold. The published Brexit "truncated" value
  (0.426) corresponds to `largest1`. With daily data and short windows,
  always compare both columns.
* **Windows are anchored, not trailing.** "1M" runs from the valuation
  date to the event date; observations are calendar-daily including
  weekends.

## The clock over time

The most instructive object is not the point estimate but the *path* of
the clock — when did the information actually arrive?

```{r}
path_gb <- event_clock_path(ep_gb, from = as.Date("2016-05-24"))
plot_clock(path_gb)
plot_clock_vs_calendar(path_gb)
```

A path below the 45-degree line means information is back-loaded — it
waits for the deadline; jumps mark days when the market learned a lot
(polls, court rulings, debates).

## Real time versus ex post

Ex post, we know how much information arrived while an option was alive.
In real time we do not; the paper's benchmark annualizes the trailing
40-observation realized variation and scales it to the horizon:

```{r}
event_clock_forecast(ep_us,
  at = as.Date("2016-10-10"),
  horizon = c(`1W` = 7, `2W` = 14)
)
```

This reproduces the published real-time values of 0.068 (1W) and 0.135
(2W) for the U.S. election — the market expected far more learning than
the 0.015/0.046 that actually materialized before election week.

## Why levels do not matter: the wedge property

On the paper's valuation date, the raw betting-quote probability of
*Leave* is $q = `r round(ep_gb$q[ep_gb$time == as.Date("2016-05-24")], 3)`$,
while the paper works with a USD-state-price-converted $q = 0.195$. The
levels differ — but the clock is identical, because an (approximately)
constant logit wedge has zero quadratic variation:

```{r}
shifted <- brexit2016
shifted$q_leave <- ec_ilogit(ec_logit(brexit2016$q_leave) + 0.5)
ep_shift <- as_event_prices(shifted, time = "date", price = "q_leave")

c(
  original = event_clock(ep_gb, from = as.Date("2016-05-24"),
                         methods = "rv")$A,
  shifted = event_clock(ep_shift, from = as.Date("2016-05-24"),
                        methods = "rv")$A
)
```

## The formula book

Once $q$ and $A$ are measured, a family of closed-form objects follows
from the exact logistic-normal transition law
$L_T = L_t + (\mathbf{1}\{J=1\} - \tfrac12) A + \sqrt{A}\,\zeta$. The
worked Brexit example ($q = 0.195$, $A = 0.166$):

```{r}
ec_moments(q = 0.195, A = 0.166)

# typical revision over the window: about 5 probability points
ec_revision(q = 0.195, A = 0.166)

# probability of ending above 50% by resolution
ec_exceedance(0.5, q = 0.195, A = 0.166)

# clock time needed to move from 19.5% to "90% sure": about 40x
# the two-week clock
ec_target_clock(q = 0.195, target = 0.9) / 0.166
```

The rule of thumb for the implied-volatility contribution of event
learning,
$\Delta\mathrm{IV} \approx [\Delta\eta\, q(1-q)]^2 A \,/\, (2\sigma(T-t))$,
reproduces the paper's headline decomposition (in annualized IV
percentage points, using the tenor-specific no-learning ATM volatility
and the event exposures $\Delta\eta$ estimated in the paper):

```{r}
c(
  brexit_2w = 100 * ec_iv_rule(deta = -0.012, q = 0.195, A = 0.166,
                               sigma = 0.10590, tenor = 14 / 365),
  us_2w     = 100 * ec_iv_rule(deta = -0.099, q = 0.172, A = 0.046,
                               sigma = 0.19326, tenor = 14 / 365)
)
```

Learning was priced at well under a tenth of a volatility point for
Brexit but an order of magnitude more for the U.S. election — the
$(\Delta\eta)^2$ lever dominates.

## The event clock on live Polymarket data

The package connects to the public, keyless Polymarket APIs. The shipped
`polymarket2024` dataset was downloaded with exactly this code:

```{r, eval = FALSE}
# 1. find the event and the "Yes" token
pm_search("presidential election winner 2024")
mkts <- pm_markets("presidential-election-winner-2024")
tok <- mkts$token_id[grepl("Trump", mkts$question) & mkts$outcome == "Yes"]

# 2. pull the hourly price history (chunked automatically)
ep24 <- pm_prices(tok, from = "2024-06-01", to = "2024-11-06",
                  market_id = "Polymarket: Trump wins 2024",
                  event_date = as.POSIXct("2024-11-05", tz = "UTC"))
```

```{r}
data(polymarket2024)
ep24 <- as_event_prices(polymarket2024,
  market_id = "Polymarket: Trump wins 2024",
  event_date = as.POSIXct("2024-11-05", tz = "UTC")
)
plot_clock(event_clock_path(ep24))
```

With intraday data, mind the sampling frequency: microstructure noise
(bid-ask bounce on a coarse tick grid) inflates measured variation.
Collapsing to one daily snapshot (last price at or before 16:00 New York
time) is a robust default:

```{r}
c(
  hourly = event_clock(ep24, methods = "rv")$A,
  daily  = event_clock(pm_daily(ep24), methods = "rv")$A
)
```

## Simulating clock-consistent paths

For teaching and testing, `ec_simulate_path()` composes the exact
transition law into full probability paths whose realized clock matches
the input $A$ and whose probabilities are martingales by construction:

```{r}
set.seed(1)
paths <- ec_simulate_path(n_paths = 20, n_steps = 200, q = 0.3, A = 1)

library(ggplot2)
ggplot(paths, aes(t_frac, q, group = path,
                  color = factor(J))) +
  geom_line(alpha = 0.7) +
  scale_color_manual(values = c(`0` = "grey60", `1` = "#2c3e50"),
                     name = "Outcome J") +
  labs(x = "Fraction of clock time elapsed", y = expression(q[t])) +
  theme_minimal(base_size = 12)
```

## Citation

If you use the package or the event-clock methodology, please cite:

> Hanke, M., Schadner, W., Stöckl, S., and Weissensteiner, A. (2026).
> *Learning Before Scheduled Events: Prediction Markets, State Prices,
> and Option Valuation.* Working Paper.

See `citation("eventclock")`.
