---
title: "Event betas: assets on the event clock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event betas: assets on the event clock}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(eventclock)
```

## From the clock to asset prices

The event clock answers "how much was learned, and when". The asset
side asks: "who cares?" To first order, an asset with outcome-conditional
mean multipliers \(\eta_1, \eta_2\) responds to probability news as
\[ r_t \;\approx\; \Delta\eta \,\Delta q_t, \qquad
   \Delta\eta = \eta_1 - \eta_2, \]
so a regression of returns on probability innovations recovers the
*event exposure* \(\widehat{\Delta\eta}\) — that is `event_beta()`.

**What returns can and cannot identify.** Returns pin down only the
spread \(\Delta\eta\), not the levels \(\eta_1, \eta_2\) separately, and
not the outcome-conditional dispersions: those require event-spanning
option smiles. Two consequences:

* the levels reported by `event_beta()` come from the risk-neutral
  adding-up constraint \(q\eta_1 + (1-q)\eta_2 = 1\) — they are
  model-implied, not independently identified;
* the loading test \(\beta = 1\) is meaningful only against an
  *externally* measured exposure (option-implied, or from an
  independent sample); regressing and testing against the same
  \(\widehat{\Delta\eta}\) would be circular. Supply it via the `deta`
  argument when you have one.

## DJT and the 2024 election

The most exposed listed asset to the 2024 U.S. presidential election
was Trump Media & Technology Group. Both ingredients ship with the
package:

```{r}
data(djt2024)
data(polymarket2024)
ep <- pm_daily(as_event_prices(polymarket2024,
  market_id = "Polymarket: Trump wins 2024",
  event_date = as.POSIXct("2024-11-05", tz = "UTC")
))

eb <- event_beta(djt2024, ep)
eb
```

The exposure is large — a ten-point move in the win probability moves
the stock by roughly \(0.1 \times \widehat{\Delta\eta} \times 100\)
percent — and highly significant despite the stock's enormous
idiosyncratic (meme) volatility, which keeps the \(R^2\) modest. The
regression \(R^2\) *is* the realized variance share of event news over
the sample.

The assassination-attempt weekend makes the mechanism visible in a
single observation: the win probability jumped by about 9 points and
the stock opened `r round(100 * diff(log(djt2024$adjusted))[which(djt2024$date == as.Date("2024-07-15")) - 1], 0)`% higher on Monday, July 15.

```{r}
plot(ep) +
  ggplot2::geom_vline(xintercept = as.Date("2024-07-13"),
                      linetype = 3, color = "grey40")
```

## Is the event first-order for option prices? The relevance screen

Given an exposure, the *sufficient statistic* for whether event
learning matters for an asset's option prices is the ratio of learning
variance to no-learning variance, \(\rho = [\Delta\eta\,
q(1-q)]^2 A / (\sigma^2 T)\):

```{r}
# ingredients: measured clock, measured exposure, the asset's own vol
A_2w <- event_clock(ep, from = as.Date("2024-10-22"),
                    to = as.Date("2024-11-05"), methods = "rv")$A
sigma_djt <- sd(diff(log(djt2024$adjusted))) * sqrt(252)
q_pre <- ep$q[ep$time == as.Date("2024-10-22")]

rho <- ec_relevance(deta = eb$deta_hat, q = q_pre, A = A_2w,
                    sigma = sigma_djt, tenor = 14 / 365)
c(rho = rho, variance_share = rho / (1 + rho),
  iv_rule_pp = 100 * sigma_djt * rho / 2)
```

Compare this with an FX pair around the same election: with
\(|\Delta\eta|\) two orders of magnitude smaller, \(\rho\) collapses to
rounding-error size — the four-lever anatomy of the rule of thumb
(exposure squared, movability squared, clock, dilution) decides who
cares about the event.

## One pipeline, many state prices

Any traded event state price feeds the same \(q \to L \to A\) machinery.
Two more converters ship with the package.

**Fed funds futures.** The 30-day fed funds future for a meeting month
settles on the monthly average rate; the standard extraction turns its
price into a meeting-implied move probability:

```{r}
data(fomc_meetings)
subset(fomc_meetings, year == 2024)[5:8, ]

# decision on the 15th of a 30-day month, pre-meeting rate 5.33%,
# futures at 94.79: a 25bp cut is ~96% priced
q_from_ffutures(94.79, pre_rate = 5.33,
                meeting_date = as.Date("2024-09-15"), step = -0.25)
```

**Merger-arb spreads.** A takeover target's price is itself an event
state price; the deal clock is the logit-QV of the implied completion
probability:

```{r}
target_path <- c(40, 41.5, 41, 44, 46, 45.5, 47, 48.5)
q <- q_from_deal_spread(target_path, offer = 50, fallback = 35)
deal <- as_event_prices(
  tibble::tibble(time = as.Date("2025-01-06") + seq_along(q) * 7, q = q),
  market_id = "Deal clock (toy example)"
)
event_clock(deal, methods = c("rv", "bipower"))
```

## Practical notes

* Match frequencies: with daily returns, collapse intraday probability
  data first (`pm_daily()`).
* Screen the probability series with `ec_validate()` and the sampling
  frequency with `ec_signature()` before interpreting exposures.
* Newey-West lags default to \(\lfloor 4(n/100)^{2/9} \rfloor\); pass
  `lags = 0` for plain heteroskedasticity-robust errors.
* Add market/factor controls via `controls =` to isolate the event
  channel from general market comovement.
